A Bidirected String Graph Model for Genome Assembly

### 2012-12-11 presentation/paper.tex
paper: updates

\newcommand{\LengthVar}{L}
\newcommand\Base[1]{{\tt #1}}
\newcommand{\ReadArrowType}{latex}
\newcommand{\Figure}[1]{Figure \ref{#1}}
\newcommand{\Subfigure}[1]{Figure \ref{#1}}
\newcommand{\Section}[1]{Section \ref{#1}}
\newcommand{\Subsection}[1]{Subsection \ref{#1}}
\newcommand{\Algorithm}[1]{Algorithm \ref{#1}}

@@

Consider the the string graph fragment shown in
\Subfigure{subfig:bidiedge_interp_a}.  When we walk from $f \to g$, we are using
both $f$ and $g$ in forward orientation, based on the arrow heads.  But what
this really means is that we are walking from the {\em end} of read $f$ to the
{\em end} of read $g$.  In other words, traversing the edge in that direction is
associated with the addition of the sequence after the overlapped region in $g$,
up to the end of the $g$.

If we instead walk $g \to f$, we instead use both $g$ and $f$ in
reverse-complement orientation by walking from the {\em beginning} of read $g$
to the {\em beginning} of read $f$.  In other words, traversing the edge in that
direction is associated with the addition of the sequence before the overlapped
region in $f$, up to the beginning of $f$, and this sequence must be
reverse-complemented because it is being used opposite its original direction.

These interpretations can be generalized to the other two types of bidirected
edges, and this is described in slightly more detail in \cite{Myers2005}.
Furthermore, these interpretations imply that the actual sequence spelled out by
a walk in the bidirected graph may include implied (non-labeled) sequence at the
start (if we start at the end of a read) and/or the end (if we end at the
beginning of a read).

\begin{figure}
	\begin{center}
		\begin{subfigure}[b]{0.48\textwidth}
			\begin{tikzpicture}[>=triangle 45,scale=1.5]
				\node[circle,fill=blue!20,draw=black] (f) at (0, 0) {f};
				\node[circle,fill=blue!20,draw=black] (g) at (2.0, 0) {g};
				\draw[>->,style=very thick] (f) edge (g);
				\draw (1.00, 0.1) node[anchor=south] {\small \tt TTAGAAGTA};
				\draw (1.00, -0.1) node[anchor=north] {
						\rotatebox{180}{\small \tt
						TGTGTAATC}};
				\draw[->] (0.5, 0.5) ..
					controls (0.7, 0.7) and (1.3, 0.7) ..
					(1.5, 0.5);
				\draw[->] (1.5, -0.5) ..
					controls (1.3, -0.7) and (0.7, -0.7) ..
					(0.5, -0.5);
			\end{tikzpicture}
			\caption{A bidirected string graph fragment}
			\label{subfig:bidiedge_interp_a}
		\end{subfigure}
		\begin{subfigure}[b]{0.48\textwidth}
			\begin{tikzpicture}[>=\ReadArrowType,scale=0.8]
				\draw[->,very thick] (0, 0) node[anchor=east] {$f.B$}
						-> (4, 0) node[anchor=west] {$f.E$};
				\draw[->,very thick] (2.09, -0.9) node[anchor=east] {$g.B$}
						-> (6.09, -0.9) node[anchor=west] {$g.E$};

				\draw (0, -0.1) node[anchor=south west]
					{\hspace*{-1.4mm}{\small \tt
						%GATTACACAGTGTAGTT
						GATTACACA\hspace*{0.26cm}$a_1 \dots a_n$
					}};

				\draw (0, 0.1) node[anchor=north west]
					{\hspace*{-2.6mm}
						\rotatebox{180}{{\small \tt TGTGTAATC}}};

@@

				\draw (2.09, -0.9) node[anchor=north west]
					{\hspace*{12.48mm}
						\rotatebox{180}{{\small \tt TACTTCTAA}}};

@@

In \Subsection{subsec:graph_overview}, the possibility of using a directed graph
to model the assembly of the reads was rejected because it did not take into
account the double-strandedness of DNA.  However, \Figure{fig:bidiedge_interp}
suggests an alternate construction of the graph that does indeed use directed
edges.  Instead of having each vertex be a read, have each vertex be an end of a
read.  The advantage to this approach is that the problem can be expressed using
a directed graph instead a bidirected graph, which can be helpful for the
implementation.  The disadvantage is that two vertices are required for each
read, they may be widely separated in the graph.

@@

The contained reads cannot, however be fully discarded, as they are needed for a
later step (see \Section{sec:mapping_contained_reads}).

\subsection{Implementation}

As described in \Subsection{subsec:removing_contained_reads}, before we even
build the graph, contained reads need to be removed.  This is implemented by the
program \ProgramName{remove-contained-reads}.  It takes as input two files: the
set of reads and the set of overlaps that were computed from them, and produces
as output three files: the set of uncontained reads, the set of overlaps
excluding those that involve contained reads, and an array that maps from the
original read indices to the new read indices.

@@

The string graphs are stored in a binary format.  An additional program,
\ProgramName{print-string-graph}, was implemented that can print a string graph
in plain-text format, or print a string graph in the DOT language for input to
the graph-drawing software {\it GraphViz}\cite{GraphVizPublication}, or print
some other information about a string graph.

@@

\section{Simplifying the graph}

@@

Before trying to find walks through the string graph, it is very helpful to
simplify the graph is much as possible.  Very commonly, given three adjacent
reads $f$, $g$, and $h$, $f$ will overlap $h$ as well as $g$.  This is redundant
because we can, equivalently, walk $f$ to $g$ to $h$ rather than directly
walking from $f$ to $h$.  \Figure{fig:transitive} shows an example of this.

@@

\Algorithm{alg:transitive_reduction} produces pseudocode for a transitive
reduction algorithm.  Essentially, the algorithm is to through each vertex $v$
and examine vertices up to 2 edges away from $v$ to identify all transitive
edges leaving $v$.  These edges are then marked for removal, and removed after
all vertices in the graph have been considered.  Some optimizations are
possible; for example, the lengths of the sequences the edges are labeled with
can be taken into account to allow terminating the search of a vertex $v$'s
neighborhood before all possible walks of length 2 leaving $v$ have been
considered.

Transitive reduction can be performed on either the bidirected or directed
representations of the string graph, although for simplicity here it is
described on the directed graph.

@@

Collapsing unbranched paths is implemented as the program
\ProgramName{collapsed-unbranched-paths}.  It takes as input a file containing a
string graph and produces as output a file containing the string graph with
unbranched paths collapsed.  Vertices are currently re-numbered; so, for
example, vertex 10 after collapsing unbranched paths does not necessarily refer
to the 10th read in the list of uncontained reads from which the graph was
constructed.

@@

\section{Examples of graph construction and simplification}

At this point, the assembler has been given a set of reads, has computed exact
overlaps from them, removed contained reads and all overlaps involving them,
then built a bidirected string graph that models the ways (via walks through the
graph) in which the reads can be assembled together when the overlaps are used
as evidence, then simplifies this graph through transitive reduction and
collapsing unbranched paths.  Usually, the graph at this point will be much
simpler than the original graph, although it may still be very complicated,
depending on the genome and the data.

@@

A completely random genome is very unlikely to have repeats of significant
length compared to the minimum overlap length.  In this case, transitive
reduction is expected to leave the entire graph as a single unbranched path,
provided the original genome was a single, non-circular string, and every base
pair of it was included in at least one read.  Then, collapsing unbranched paths
will collapse the entire bidirected graph into two vertices connected by a
single edge.  Finding a walk through the edges of this graph is trivial and will
produce an assembly of the entire genome.  \Figure{fig:random_example} shows the
bidirected string graph that might result from such a genome and data.

If we still assume a genome with no repeats of length greater than the overlap
length, but relax the assumption that the genome is a single, non-circular
string from which every base pair was included in a read, the graph will instead
reduce and collapse to a set of components.  Components corresponding to
non-circular sequence will contain two vertices with an edge between, while
components corresponding to circular sequence will contain one vertex with an
edge connecting it to itself (a loop).  In this case (and as will be elaborated
on later), finding a walk through the bidirected string graph must be
generalized to allow walking each component separately; but again for this
particular case of a random genome, each such walk is still trivial, and the
assembly will be optimal as long as the assembler does not decide to traverse
any of the loops more than one time (this will also be elaborated on more
later).

\begin{figure}
	\begin{center}
		\begin{tikzpicture}
			{\small
			\begin{tikzpicture}[>=triangle 45]
				\node[circle,fill=blue!20,draw=black] (A) at (0.0, 0) {A};
				\node[circle,fill=blue!20,draw=black] (B) at (3.0, 0) {B};
				\draw[>->,very thick] (A) edge node[anchor=north] {4639100} (B);
			\end{tikzpicture} }
		\end{tikzpicture}
		\caption{
			Bidirected string graph for reads uniformly sampled from a 4.6 Mbp
			random genome after transitive reduction and path collapsing.  (Edge
			labeled with length of sequence, with the length being the same in both
			directions)}
		\label{fig:random_example}
	\end{center}
\end{figure}

@@

\begin{figure}
	\begin{center}
		\includegraphics[width=0.6\textwidth]{E_coli.pdf}
		\caption{Transitively-reduced, collapsed bidirected string graph for
		1000 bp perfect reads uniformly sampled every 100 bp from random strands
		of the non-circular version of the 4,639,675 bp genome of {\it E. coli}
		(a common bacterium) using a minimum overlap length of 250 bp.  This
		graph was produced by the code that has been discussed, but the actual
		drawing of the picture was done by {\it
		GraphViz}\cite{GraphVizPublication}.  The 46,387 reads
		that were provided as input to the assembler were generated by a Perl
		script.  The bottleneck of the algorithm up to this point is
		\ProgramName{compute-overlaps}, which took about 10 minutes and used a
		maximum of about 1 GiB of memory to find all 336,227 overlaps from these
		reads, seeding on 128-mers.  The running time and memory usage of
		\ProgramName{compute-overlaps} would need to improved further to handle
		larger genomes.  (Also it still does not handle inexact overlaps, which
		are not considered in this paper.)}
		\label{fig:E_coli}
	\end{center}
\end{figure}

@@

As \Figure{fig:E_coli} shows, the bidirected string graph may still be very
complicated even after transitive reduction and traversal.  In even worse
examples, the graph could easily have thousands of vertices and edges.
Therefore, we need to figure out how to walk through the graph to produce the
best assembly.

@@

However, there are multiple problems with the Eulerian path approach.  First,
the graph may not be Eulerian.  This can happen if the graph is not strongly
connected, not balanced (indegree always equal to out

This implies we really would need to solve the
bidirected equivalent of the Chinese Postman Problem instead.  Even then, the
graph may not be fully connected, which would require that the final walk take
the form of multiple discontinuous walks.  An equally severe problem is that
some overlaps may be spurious, meaning that the two overlapping reads do not
actually form a valid genomic sequence when combined.  Spurious overlaps can be
caused by genomic repeats.  Therefore, we need to allow for the possibility that
a given edge in the bidirected graph is omitted from the assembly if it seems
like a good idea to do so.

The way these problems can be solved is to (1) place a lower bound and an upper
bound on the number of times each edge may be traversed, then (2) formulate and
solve a minimum-cost network flow problem that finds the minimum total traversal
count of edges in the graph in order for every vertex to be balanced (i.e. have
the same indegree as out degree), then (3) find a generalized Eulerian tour
through the graph, traversing each edge the exact number of times that was
calculated in (2) and also allowing for the tour be discontinuous.

@@

We would like to use information from the data to restrict the number of times
each edge may be traversed.  This can be done using a calculation called the
\KeyTerm{A-statistic}, or \KeyTerm{arrival-rate statistic}, which is the
logarithm of the probability that an edge represents unique sequence rather than
duplicate sequence.  The A-statistic $A(e)$ of an edge $e$ in a bidirected
string graph is:

>>>>

\newcommand{\LengthVar}{L}
\newcommand\Base[1]{{\tt #1}}
\newcommand{\ReadArrowType}{latex}
\newcommand{\Figure}[1]{Figure \ref{#1}}
\newcommand{\Subfigure}[1]{Figure \ref{#1}}
\newcommand{\Section}[1]{Section \ref{#1}}
\newcommand{\Subsection}[1]{Section \ref{#1}}
\newcommand{\Algorithm}[1]{Algorithm \ref{#1}}

@@

Consider the the string graph fragment shown in
\Subfigure{subfig:bidiedge_interp_a}.  When we walk $f \to g$, both $f$ and $g$
are used in forward orientation, as indicated by the directions of the adjacent
arrow heads.  But what this really means is that we are walking from the {\em
end} of read $f$ to the {\em end} of read $g$.  In other words, traversing the
edge in that direction is associated with the addition of the sequence after the
overlapped region in $g$, up to the end of the $g$.

If we instead walk $g \to f$, both $g$ and $f$ are used reverse-complement
orientation by walking from the {\em beginning} of read $g$ to the {\em
beginning} of read $f$.  In other words, traversing the edge in that direction
is associated with the addition of the sequence before the overlapped region in
$f$, up to the beginning of $f$, and this sequence must be reverse-complemented
because it is being used opposite its original direction.

These interpretations can be generalized to the other two types of bidirected
edges, and this is described in slightly more detail in \cite{Myers2005}.
Furthermore, these interpretations imply that the actual sequence spelled out by
a walk in the bidirected graph may include implied (non-labeled) sequence at the
start (if we start at the end of a read) and/or the end (if we end at the
beginning of a read) of the walk.

\begin{figure}
	\begin{center}
		\begin{subfigure}[b]{0.48\textwidth}
			\begin{tikzpicture}[>=triangle 45,scale=1.5]
				\node[circle,fill=blue!20,draw=black] (f) at (0, 0) {f};
				\node[circle,fill=blue!20,draw=black] (g) at (2.0, 0) {g};
				\draw[>->,style=very thick] (f) edge (g);
				\draw (1.00, 0.1) node[anchor=south] {\small \tt TTAGAAGTA};
				\draw (1.00, -0.1) node[anchor=north] {
						\rotatebox{180}{\small \tt
						TGTGTAATC}};
				\draw[->] (0.5, 0.5) ..
					controls (0.7, 0.7) and (1.3, 0.7) ..
					(1.5, 0.5);
				\draw[->] (1.5, -0.5) ..
					controls (1.3, -0.7) and (0.7, -0.7) ..
					(0.5, -0.5);
			\end{tikzpicture}
			\caption{A bidirected string graph fragment}
			\label{subfig:bidiedge_interp_a}
		\end{subfigure}
		\begin{subfigure}[b]{0.48\textwidth}
			\begin{tikzpicture}[>=\ReadArrowType,scale=0.8]
				\draw[->,very thick] (-0.2, 0) node[anchor=east] {$f.B$}
						-> (4, 0) node[anchor=west] {$f.E$};
				\draw[->,very thick] (2.09, -0.9) node[anchor=east] {$g.B$}
						-> (6.29, -0.9) node[anchor=west] {$g.E$};

				\draw (-0.2, -0.1) node[anchor=south west]
					{\hspace*{-1.5mm}{\small \tt
						%GATTACACAGTGTAGTT
						GATTACACA\hspace*{1.5mm}$a_1 \dots a_n$
					}};

				\draw (-0.2, 0.1) node[anchor=north west]
					{\hspace*{-2.6mm}
						\rotatebox{180}{{\small \tt TGTGTAATC}}};

@@

				\draw (2.09, -0.9) node[anchor=north west]
					{\hspace*{12.88mm}
						\rotatebox{180}{{\small \tt TACTTCTAA}}};

@@

In \Subsection{subsec:graph_overview}, the possibility of using a directed graph
to model the assembly of the reads was rejected because it did not take into
account the double-strandedness of DNA.  However, \Figure{fig:bidiedge_interp}
suggests an alternate construction of the graph that does indeed use directed
edges.  Instead of having each vertex be a read, have each vertex be an end of a
read.  The advantage to this approach is that the problem can be expressed using
a directed graph instead a bidirected graph, which can be helpful for the
implementation.  The disadvantage is that there are two completely separate
vertices for each read.

@@

This implies that a vertex in the bidirected string graph may actually
correspond to more than one read.

The contained reads cannot be completely discarded because they are needed for a
later step (see \Section{sec:mapping_contained_reads}).

\subsection{Implementation of assembly graph construction}

As described in \Subsection{subsec:removing_contained_reads}, before the
assembly graph is even built, contained reads need to be removed.  This is
implemented by the program \ProgramName{remove-contained-reads}.  It takes as
input two files: the set of reads and the set of overlaps that were computed
from them, and produces as output three files: the set of uncontained reads, the
set of overlaps excluding those that involve contained reads, and an array that
maps from the original read indices to the new read indices.

@@

The string graphs are stored in a binary format.  An additional program,
\ProgramName{print-string-graph}, was implemented that can print a string graph
in plain-text format, print a string graph in the DOT language for input to the
graph-drawing software {\it GraphViz}\cite{GraphVizPublication}, or print some
other information about a string graph.

@@

\section{Simplifying the assembly graph}

@@

Before trying to find walks through the string graph, it is very helpful to
simplify the graph is much as possible.  Very commonly, given three adjacent
reads $f$, $g$, and $h$, $f$ will overlap $h$ as well as $g$.  This is redundant
because we can, equivalently, walk $f \to g \to h$ rather than directly walking
from $f \to h$.  \Figure{fig:transitive} shows an example of this.

@@

\Algorithm{alg:transitive_reduction} provides pseudocode for a transitive
reduction algorithm similar to that described in \cite{Myers2005}.  Essentially,
the algorithm is to through each vertex $v$ and examine vertices up to 2 edges
away from $v$ to identify all transitive edges leaving $v$.  These edges are
then marked for removal, and they are removed after all vertices in the graph
have been considered.  Some optimizations are possible; for example, the lengths
of the sequences the edges are labeled with can be taken into account to allow
terminating the search of $v$'s neighborhood before all possible walks of length
2 starting at $v$ have been considered.

Transitive reduction can be performed on either the bidirected or directed
representations of the string graph, although for simplicity
\Algorithm{alg:transitive_reduction} implies the directed representation.

@@

Collapsing unbranched paths is implemented as the program
\ProgramName{collapse-unbranched-paths}.  It takes as input a file containing a
string graph and produces as output a file containing the string graph with
unbranched paths collapsed.  Vertices are currently re-numbered; so, for
example, vertex 10 after collapsing unbranched paths does not necessarily refer
to the 10th read in the list of uncontained reads from which the graph was
constructed.

@@

\section{Examples of assembly graph construction and simplification}

At this point, the assembler has been given a set of reads, has computed exact
overlaps from them, has removed contained reads and all overlaps involving them,
has built a bidirected string graph that models the ways in which the reads can
be assembled together when the overlaps are used as evidence, and has simplified
this graph through transitive reduction and collapsing unbranched paths.
Usually, the graph at this point will be much simpler than the original graph,
although it may still be very complicated, depending on the genome and the data.

@@

A completely random genome is very unlikely to have repeats of significant
length compared to the minimum overlap length.  In this case, transitive
reduction is expected to leave the entire graph as a single unbranched path,
provided the original genome was a single, non-circular string and every base
pair of it was included in at least one read.  Then, collapsing unbranched paths
will collapse the entire bidirected graph into two vertices connected by a
single edge.  Finding a walk through the edges of this graph is trivial and will
produce an assembly of the entire genome.  \Figure{fig:random_example} shows the
bidirected string graph that might result from such a genome and data.

If we still assume a genome with no repeats of length greater than the overlap
length, but relax the assumption that the genome is a single, non-circular
string from which every base pair was included in a read, the graph will instead
reduce and collapse to a set of components.  Components corresponding to
non-circular sequence will contain two vertices with an edge between, while
components corresponding to circular sequence will contain one vertex with an
edge connecting it to itself (a loop).  In this case (and as will be elaborated
on later), finding a walk through the bidirected string graph must be
generalized to allow walking each component separately; but again for this
particular case of a random genome, each such walk is still trivial, and the
assembly will be optimal as long as the assembler does not decide to traverse
any of the loops more than one time (this will also be elaborated on more
later), since there are no repeats of substantial length in a hypothetical
random genome.

\begin{figure}
	\begin{center}
		\vspace{1cm}
		\begin{tikzpicture}
			{\small
			\begin{tikzpicture}[>=triangle 45]
				\node[circle,fill=blue!20,draw=black] (A) at (0.0, 0) {A};
				\node[circle,fill=blue!20,draw=black] (B) at (3.0, 0) {B};
				\draw[>->,very thick] (A) edge node[anchor=north] {4639100} (B);
			\end{tikzpicture} }
		\end{tikzpicture}
		\caption{
			Bidirected string graph for reads uniformly sampled from a 4.6 Mbp
			random genome after transitive reduction and path collapsing.  (Edge
			labeled with length of sequence, with the length being the same in both
			directions)}
		\label{fig:random_example}
	\end{center}
\end{figure}

@@

\begin{figure}
	\begin{center}
		\includegraphics[width=1.0\textwidth]{E_coli.pdf}
		\caption{Transitively-reduced, collapsed bidirected string graph for
		1000 bp perfect reads uniformly sampled every 100 bp from random strands
		of the non-circular version of the 4,639,675 bp genome of {\it E. coli}
		(a common bacterium) using a minimum overlap length of 250 bp.  This
		graph was produced by the code that has been discussed, but the actual
		drawing of the picture was done by {\it
		GraphViz}\cite{GraphVizPublication}.  The 46,387 reads
		that were provided as input to the assembler were generated by a Perl
		script.  The bottleneck of the algorithm up to this point is
		\ProgramName{compute-overlaps}, which took about 10 minutes and used a
		maximum of about 1 GiB of memory to find all 336,227 overlaps from these
		reads, seeding on 128-mers.  The running time and memory usage of
		\ProgramName{compute-overlaps} would need to improved further to handle
		larger genomes.}
		\label{fig:E_coli}
	\end{center}
\end{figure}

@@

As \Figure{fig:E_coli} shows, the bidirected string graph may still be very
complicated even after transitive reduction and traversal.  In even worse
examples, the graph could easily have thousands of vertices and edges, and it
may not even be connected.  Therefore, we need to figure out how to walk through
the graph to produce the best assembly.

@@

However, there are multiple problems with the Eulerian path approach.  First,
the graph may not be Eulerian.  This could happen if the graph is not balanced.
This implies we really would need to solve the bidirected equivalent of the
Chinese Postman Problem instead.\footnote{The Chinese Postman problem on a
bidirected graph is formalized in \cite{Medvedev2009}.}  Even then, the graph
may not be fully connected, which would require that the final walk take the
form of multiple discontinuous walks.  An equally severe problem is that some
overlaps may be spurious, meaning that the two overlapping reads do not actually
form a valid genomic sequence when combined.  Spurious overlaps can be caused by
genomic repeats.  Therefore, we need to allow for the possibility that an edge
in the bidirected graph can be omitted from the walk if it seems like a good
idea to do so.

The way these problems can be solved is to (1) place a lower bound and an upper
bound on the number of times each edge may be traversed, then (2) formulate and
solve a minimum-cost network flow problem that finds the minimum total traversal
count of edges in the graph in order for every vertex to be balanced (i.e. have
the same in-degree as out-degree), then (3) find a generalized Eulerian tour
through the bidirected graph, traversing each edge the exact number of times
that was calculated in (2) and also allowing for the tour be discontinuous.

@@

We would like to use hints from the original data to restrict the number of
times each bidirected edge may be traversed.  This can be done using a
calculation called the \KeyTerm{A-statistic}, or \KeyTerm{arrival-rate
statistic}, which is the logarithm of the probability that an edge represents
unique sequence rather than duplicate sequence\cite{Myers2005}.  The A-statistic
$A(e)$ of an edge $e$ in a bidirected string graph is:
